### 2017-05-05 manuscript.rmd
removed hyphen in title
added space before citations

@@

For each internal node in each gene tree, we calculated the phylogenetic independent contrast[@Felsenstein:1985ua] (PIC) of Tau. This is the difference in values of Tau for descendant nodes scaled by the expected variance, which is largely determined by the lengths of the two branches that connect the node to its two descendants (Figure 1*B*). These contrasts were then annotated by whether each is made across a speciation or duplication event. The original description of independent contrasts [@Felsenstein:1985ua] focused on assessing covariance between changes in two traits. Our use of contrasts is a bit different-- we look for differences in evolutionary changes of one trait (differential expression) between two categories of nodes (speciation and duplication).

@@

In order to better understand why our phylogenetic analysis supports a different conclusion (*i.e.*, no support for the ortholog conjecture) than the published analysis of KMRR [@KryuchkovaMostacci:2016iw] (*i.e.*, strong support for the ortholog conjecture), we first checked to make sure we could reproduce their result based on pairwise analyses. This is important since we are only looking at a subset of the data they considered, the Brawand *et al.* 2011 [@Brawand:2011du] dataset for gene trees that could be successfully time calibrated. In their Figure 1[@KryuchkovaMostacci:2016iw], they present a higher Tau correlation coefficient between ortholog pairs than paralog pairs. We find the same here, with correlation coefficients of `r signif( ortholog_r, 2)` for orthologs and `r signif( paralog_r, 2)` for paralogs.

@@

The KMRR study did investigate the impact of node age on correlation, but in a different way. In their Figure 2[@KryuchkovaMostacci:2016iw], they grouped orthologs and paralogs according to the ENSEMBL node name of their most recent common ancestor, and plotted the correlation of Tau for each of these groups by the node age. They found that across the investigated range of node ages, ortholog pairs have higher Tau correlation than paralogs. We confirmed that we can replicate this result (Figure 2*A*). There are several difficulties with interpreting this plot, though. First, it does not just reflect the evolutionary processes that generated the data, it is also impacted by the phylogenies along which these processes acted. The expected covariance of traits that evolve under neutral processes is in fact defined by the phylogeny [@Grafen:1989um]. Second, the correlation for each group is based on multiple non-independent pairwise comparisons.

@@

Levin *et al.* [@Levin:2016bp] arrived at this conclusion by making multiple pairwise comparisons of ortholog expression data sampled throughout the course of embryonic development. For each species pair, they identified the orthologs shared by these species. This list of shared genes was different from species pair to species pair. They characterized each of these orthologs in each species as having expression that peaks in early, mid, or late temporal phase of development. They then calculated a similarity score for each temporal phase for each species pair based on the fraction of genes that exhibited the same patterns in each species. The distributions of similarity scores are plotted in their Figure 4d[@Levin:2016bp], and their Kolmogorov–Smirnov (KS) tests indicated that the early distribution and late distribution were each significantly different from mid distribution (P < $10^{-6}$ and P < $10^{-12}$, respectively). This is the support they presented for the inverse hourglass model.

@@

```{r cor_structure_plot, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, dev="png", dpi=600, fig.width=6.5, fig.height=6.5, fig.cap="Origin of pairwise correlation plot structure. (*A*) Each speciation and duplication node in the COMPARA trees is annotated with a clade name, and the age of that clade (shown on the x axis) is used as the age of the pairwise comparisons (Figure 2 *A, C, E*, KMRR[@KryuchkovaMostacci:2016iw] Figure 2). The clade ages have a clear correspondence to the speciation node ages (red). Duplication (blue) ages for nodes assigned to the same clade, however, have large variation in calibrated node age (y axis). This means that summaries of expression differences across duplication events represent evolutionary changes that occurred over a wide range of time scales. (*B*) When duplication nodes whose calibrated node age deviates more than 10% from their annotated clade age are removed from the data simulated under the null distribution, duplication comparisons no longer have a uniformly lower expression correlation than speciation comparisons."}

@@

KMRR[@KryuchkovaMostacci:2016iw] Figure 3 differs from their other analyses in that it presents independent changes in expression between triplets of genes. Each triplet has two paralogs that arose from a duplication event and one unduplicated homolog. They find that there is a tendency for the paralog with the lowest expression to have the highest Tau. From this they conclude that following duplication there is a common trajectory of evolutionary change, where one paralog evolves to have lower expression and also become more tissue specific. There is, however, a negative relationship between Tau and maximum expression across all genes (*SI Appendix* Figure 7, $r^2=`r signif( tau_v_max_lm$r.squared, 3 )`$, $slope=`r signif( tau_v_max_lm$coefficients[2,1], 3 )`$, $p=`r signif( tau_v_max_lm$coefficients[2,4], 3 )`$). A simpler explanation for their plot is that it reflects the global tendency for Tau to decrease with maximum expression. The null expectation for any two genes sampled at random is for one to have higher maximum expression and lower Tau, and the other to have lower maximum expression and higher Tau. This global pattern could be due to both biological factors and the technical details of how Tau is calculated.

>>>>

\hyphenation{analyzing}

@@

For each internal node in each gene tree, we calculated the phylogenetic independent contrast [@Felsenstein:1985ua] (PIC) of Tau. This is the difference in values of Tau for descendant nodes scaled by the expected variance, which is largely determined by the lengths of the two branches that connect the node to its two descendants (Figure 1*B*). These contrasts were then annotated by whether each is made across a speciation or duplication event. The original description of independent contrasts [@Felsenstein:1985ua] focused on assessing covariance between changes in two traits. Our use of contrasts is a bit different-- we look for differences in evolutionary changes of one trait (differential expression) between two categories of nodes (speciation and duplication).

@@

In order to better understand why our phylogenetic analysis supports a different conclusion (*i.e.*, no support for the ortholog conjecture) than the published analysis of KMRR [@KryuchkovaMostacci:2016iw] (*i.e.*, strong support for the ortholog conjecture), we first checked to make sure we could reproduce their result based on pairwise analyses. This is important since we are only looking at a subset of the data they considered, the Brawand *et al.* 2011 [@Brawand:2011du] dataset for gene trees that could be successfully time calibrated. In their Figure 1 [@KryuchkovaMostacci:2016iw], they present a higher Tau correlation coefficient between ortholog pairs than paralog pairs. We find the same here, with correlation coefficients of `r signif( ortholog_r, 2)` for orthologs and `r signif( paralog_r, 2)` for paralogs.

@@

The KMRR study did investigate the impact of node age on correlation, but in a different way. In their Figure 2 [@KryuchkovaMostacci:2016iw], they grouped orthologs and paralogs according to the ENSEMBL node name of their most recent common ancestor, and plotted the correlation of Tau for each of these groups by the node age. They found that across the investigated range of node ages, ortholog pairs have higher Tau correlation than paralogs. We confirmed that we can replicate this result (Figure 2*A*). There are several difficulties with interpreting this plot, though. First, it does not just reflect the evolutionary processes that generated the data, it is also impacted by the phylogenies along which these processes acted. The expected covariance of traits that evolve under neutral processes is in fact defined by the phylogeny [@Grafen:1989um]. Second, the correlation for each group is based on multiple non-independent pairwise comparisons.

@@

Levin *et al.* [@Levin:2016bp] arrived at this conclusion by making multiple pairwise comparisons of ortholog expression data sampled throughout the course of embryonic development. For each species pair, they identified the orthologs shared by these species. This list of shared genes was different from species pair to species pair. They characterized each of these orthologs in each species as having expression that peaks in early, mid, or late temporal phase of development. They then calculated a similarity score for each temporal phase for each species pair based on the fraction of genes that exhibited the same patterns in each species. The distributions of similarity scores are plotted in their Figure 4d [@Levin:2016bp], and their Kolmogorov–Smirnov (KS) tests indicated that the early distribution and late distribution were each significantly different from mid distribution (P < $10^{-6}$ and P < $10^{-12}$, respectively). This is the support they presented for the inverse hourglass model.

@@

```{r cor_structure_plot, cache=TRUE, echo=FALSE, warning=FALSE, message=FALSE, dev="png", dpi=600, fig.width=6.5, fig.height=6.5, fig.cap="Origin of pairwise correlation plot structure. (*A*) Each speciation and duplication node in the COMPARA trees is annotated with a clade name, and the age of that clade (shown on the x axis) is used as the age of the pairwise comparisons (Figure 2 *A, C, E*, KMRR [@KryuchkovaMostacci:2016iw] Figure 2). The clade ages have a clear correspondence to the speciation node ages (red). Duplication (blue) ages for nodes assigned to the same clade, however, have large variation in calibrated node age (y axis). This means that summaries of expression differences across duplication events represent evolutionary changes that occurred over a wide range of time scales. (*B*) When duplication nodes whose calibrated node age deviates more than 10% from their annotated clade age are removed from the data simulated under the null distribution, duplication comparisons no longer have a uniformly lower expression correlation than speciation comparisons."}

@@

KMRR [@KryuchkovaMostacci:2016iw] Figure 3 differs from their other analyses in that it presents independent changes in expression between triplets of genes. Each triplet has two paralogs that arose from a duplication event and one unduplicated homolog. They find that there is a tendency for the paralog with the lowest expression to have the highest Tau. From this they conclude that following duplication there is a common trajectory of evolutionary change, where one paralog evolves to have lower expression and also become more tissue specific. There is, however, a negative relationship between Tau and maximum expression across all genes (*SI Appendix* Figure 7, $r^2=`r signif( tau_v_max_lm$r.squared, 3 )`$, $slope=`r signif( tau_v_max_lm$coefficients[2,1], 3 )`$, $p=`r signif( tau_v_max_lm$coefficients[2,4], 3 )`$). A simpler explanation for their plot is that it reflects the global tendency for Tau to decrease with maximum expression. The null expectation for any two genes sampled at random is for one to have higher maximum expression and lower Tau, and the other to have lower maximum expression and higher Tau. This global pattern could be due to both biological factors and the technical details of how Tau is calculated.
